---
title: "customFunction"
output: html_document
---
## how to make custom function using a fork of Math with cmdstanr

1. Forking math to your own Github
I have forked Stan math (https://github.com/stan-dev/math) to my own repository (https://github.com/hyunjimoon/math).

2. Writing the function you want to add to Math in your own branch
For debug purpose, prepare some test data. We used interpolate function from `pramcma` to retrieve interpolated y values corresponding to `x_test`.

```{r setup, include=FALSE}
library(tidyverse)
library(pracma)

# Interpolate a sine wave at a few points

x = seq(0.0, 5.0, length = 5)
y = sin(x)

# Print the values so can copy them to Stan

options(digits = 10)
print(x)
print(y)

# Plot the interpolated function against the true function

tibble(x_int = seq(min(x), max(x), length = 500)) %>%
  mutate(y_true = sin(x_int),
         y_int = interp1(x, y, x_int)) %>%
  gather(which, value, y_true, y_int) %>%
  ggplot() +
  geom_line(aes(x_int, value, group = which, color = which))

# Here are some points at which we can test our function

x_test = c(1.0, 1.7, 2.1, 4.0)

# Print values at these points (to test Stan implementation)

interp1(x, y, x_test)

fd_deriv = function(x_test) {
  dx = 1e-8
  (interp1(x, y, x_test + dx) - interp1(x, y, x_test - dx)) / (2 * dx)
}

# Print gradients at these points (to test Stan implementation)
# These won't be defined everywhere for linear interpolation

fd_deriv(x_test)
```

3. Writing tests for the function you added in your own branch
Code for interpolation is as follows which goes into `stan/math/prim/fun/interp1.hpp`.
```{c++}
#ifndef STAN_MATH_PRIM_FUN_INTERP1_HPP
#define STAN_MATH_PRIM_FUN_INTERP1_HPP

using namespace std;
namespace stan {
    namespace math {
        template <typename T>
        inline std::vector<T> interp1(const std::vector<double>& xData, const std::vector<double>& yData, const std::vector<T>& xTest)
        {
            using stan::math::index_type_t;
            check_size_match("interp1", "x", xData.size(), "y", yData.size());
            check_sorted("interp1", "x", xData);
            int N = xData.size();
            std::vector<double> dydx(N);
            int M = xTest.size();
            std::vector<T> yVals(M);

            for (int i = 0; i < N - 1; i++){
                dydx[i] = (yData[i+1] -  yData[i]) / (xData[i+1] - xData[i]);
            }
            for(int j = 0; j < M; j ++){
                T x = xTest[j];
                int i= 0;
                while(x > xData[i]){
                    i++;
                }
                if(x < xData[0]){yVals[j] = yData[0];}
                else if (x > xData[N-1]){yVals[j] = yData[N-1];}
                else{yVals[j] = yData[i-1] + dydx[i-1] * (x - xData[i-1]);}
            }
            return yVals;
        }
    }
}

#endif

```

Both the value and its gradient of the function's output should be tested. The file for the former goes in `test/unit/math/prim/fun/` and to test, type `python runTest.py test/unit/math/prim/fun/interp1_test.cpp` on your command.
```{c++}
#include <gtest/gtest.h>
#include <stan/math/prim.hpp>
#include <test/unit/math/expect_near_rel.hpp>
#include <vector>

TEST(MathPrim, interp1_test) {
using stan::test::expect_near_rel;
std::vector<double> x = {0.00, 1.25, 2.50, 3.75, 5.00};
std::vector<double> y = {0.0000000000, 0.9489846194, 0.5984721441, -0.5715613187, -0.9589242747};

std::vector<double> x_test = {1.0, 1.7, 2.1, 4.0};
std::vector<double> y_ref = {0.7591876955, 0.8228001283, 0.7106361362, -0.6490339099}; // Reference values from R

std::vector<double> y_test = stan::math::interp1(x, y, x_test);

stan::test::expect_near_rel("test Interp1", y_test, y_ref);
for (auto i: y_test)
    std::cout << i << ' ';
}
```

For the gradient test, autodiff, the file goes into `test/unit/math/rev/fun/` and can be tested with the command `python test/unit/math/rev/fun/interp1_test.cpp`
```{c++}
#include <stan/math/rev.hpp>
#include <gtest/gtest.h>
#include <limits>

TEST(AgradRev, interp1_gradient) {
using stan::math::var;
using stan::math::interp1;

std::vector<double> x = { 0.1, 1.2 };
std::vector<double> y = { 1.0, -2.0 };

var x_int = 0.7;
std::vector<var> x_int_vec = { x_int };

std::vector<var> y_int_vec = interp1(x, y, x_int_vec);
var y_int = y_int_vec[0];

//var y_int = interp1(x, y, x_int_vec);

y_int.grad();

// x_int.adj() should contain dy_int/dx_int
EXPECT_FLOAT_EQ(-3.0 / 1.1, x_int.adj());
}
```

4. Forking Cmdstan/Stan to your own Github and pointing them at Math
With the followig code, custom function code at certain math branch could be cloned and rebuilt into cmdstan. 
```{r}
clone_cmdstan <- function(
  clone_dir,
  cmdstan_url = "http://github.com/stan-dev/cmdstan",
  stan_url = "http://github.com/stan-dev/stan",
  math_url = "http://github.com/hyunjimoon/math",
  cmdstan_branch = "develop",
  stan_branch = "develop",
  math_branch = "feature/interp1",
  cores = 4) {
  if (!dir.exists(clone_dir)) {
    dir.create(clone_dir)  
  }
  git2r::clone(url = cmdstan_url, branch = cmdstan_branch, local_path = file.path(clone_dir))
  git2r::clone(url = stan_url, branch = stan_branch, local_path = file.path(clone_dir, "stan"))
  git2r::clone(url = math_url, branch = math_branch, local_path = file.path(clone_dir, "stan", "lib", "stan_math"))
  cmdstanr::rebuild_cmdstan(dir = clone_dir, cores = cores)
}

```

5. Using custom Cmdstan/Stan/Math stack with cmdstanr
New function `interp1` could be tested with the following stan file.
```{stan}
functions {
  real[] interp1(real[] x, real[] y, real[] x_test);
}

transformed data {
  real x[2] = { -5.0, 10.0 };
  real y[2] = { -3.0, 0.0 };
}

parameters {
  real<lower = -0.5, upper = 10.0> x_int[1];
}

transformed parameters {
  real y_int = interp1(x, y, x_int)[1];
}

model {
  x_int ~ normal(0, 5);
}
```

Because the slope of the line here is 1/5, then y_int should have a standard normal distribution if everything is `interp1` is functioning as we have extecpted.To check this posterior intervals line up with what we would expect from a standard normal: qnorm(c(0.05, 0.25, 0.75, 0.95), 0, 1).

```{r}
library("cmdstanr")
set_cmdstan_path("customFunc_cmdstan")
mod <- cmdstan_model("interp1.stan", compile = FALSE)
modc <- mod$compile(stanc_options = list("allow-undefined" = TRUE))
fit <- mod$sample(iter_warmup = 0, iter_sampling = 10)
plot(fit$x_int)
```


The whole process is documented in [here](https://github.com/hyunjimoon/math/pull/1).